\documentclass[onefignum,onetabnum]{siamltex}
% \documentclass[final,leqno,onefignum,onetabnum]{siamltex1213}
% \usepackage[T1]{fontenc}
% \usepackage[latin9]{inputenc}
% \usepackage{float}
% \usepackage{amstext}
% \usepackage{amssymb}

% Short headings should be running head and authors last names

% \ShortHeadings{Computing the log-determinant of symmetric, diagonally dominant matrices
% in near-linear time}{Hunter, {El Alaoui} and Bayen}
% \firstpageno{1}


\usepackage[]{algorithm2e}
% \usepackage{float}
\usepackage{amstext}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{lineno}
\linenumbers
% \makeatletter

\input{preamble}

\title{Computing the log-determinant of symmetric, diagonally dominant matrices
in near-linear time}

% \author{\name Timothy Hunter \email tjhunter@eecs.berkeley.edu \\
%        \name Ahmed {El Alaoui} \email elalaoui@eecs.berkeley.edu \\
%        \name Alexandre M. Bayen \email bayen@berkeley.edu \\
%        \addr Department of Electrical Engineering and Computer Sciences\\
%        University of California\\
%        Berkeley, CA 94720-1776, USA}
\author{Timothy Hunter\footnotemark[2] \and 
Ahmed {El Alaoui}\footnotemark[2] \and
Alexandre M. Bayen\footnotemark[2]}

\footnotetext[2]{Department of Electrical Engineering and Computer Sciences, University of California, Berkeley.}
%\footnotetext[3]{\email{tjhunter@eecs.berkeley.edu}}
%\footnotetext[4]{\emai{lelalaoui@eecs.berkeley.edu}}
%\footnotetext[5]{\email{bayen@berkeley.edu}}

% \usepackage{babel}
%   \providecommand{\corollaryname}{Corollary}
%   \providecommand{\definitionname}{Definition}
%   \providecommand{\lemmaname}{Lemma}
%   \providecommand{\propositionname}{Proposition}
% \providecommand{\theoremname}{Theorem}

\begin{document}
\maketitle
% \slugger{sicomp}{xxxx}{xx}{x}{x--x}%slugger should be set to mms, siap, sicomp, sicon, sidma, sima, simax, sinum, siopt, sisc, or sirev

\begin{abstract}
We present new algorithms for computing the log-determinant of symmetric,
diagonally dominant matrices. Existing algorithms run with cubic complexity
with respect to the size of the matrix in the worst case. Our algorithm 
computes an approximation of the log-determinant in time near-linear with 
respect to the number of non-zero entries and with high probability. This 
algorithm builds upon the utra-sparsifiers introduced by Spielman and Teng for
Laplacian matrices and ultimately uses their refined versions introduced by Koutis, Miller and Peng. We also present simpler algorithms that compute upper and 
lower bounds and that may be of more immediate practical interest.
\end{abstract}

\begin{keywords}
Log-determinant.
Determinant preconditioners.
SDD matrices.
Graph sparsification.
Laplacian generalized stretch.
\end{keywords}

\begin{AMS}\end{AMS}


\pagestyle{myheadings}
\bibliographystyle{siam}
\thispagestyle{plain}
\markboth{Hunter, El Alaoui, Bayen }{Computing the log-determinant of SDD matrices
in near-linear time} 

% \editor{Editor}


\input{detsdd_1introduction}
\input{detsdd_2preconditionedlogdet}
\input{detsdd_3laplacian_problem}
\input{detsdd_31tree_preconditioner}
\input{detsdd_33ultra_preconditioner}
\input{detsdd_34stretch_bounds}
\input{detsdd_7conclusion}
\input{detsdd_8appendixa}
\input{detsdd_9appendixb}


% \bibliographystyle{plain}
\bibliography{2012_sparse_info}
% \bibliography{2012_sparse_info,siam}

\end{document}
